---
title: "Diamond sizes"
date: 2016-08-25
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(ggplot2)
library(dplyr)

smaller <- diamonds %>% 
  filter(carat <= 2.5)
```

We have data about `r nrow(diamonds)` diamonds. Only 
`r nrow(diamonds) - nrow(smaller)` are larger than
2.5 carats. The distribution of the remainder is shown
below:

```{r}
smaller %>% 
  ggplot(aes(carat)) + 
  geom_freqpoly(binwidth = 0.01)
```

## Diamond size by cut

```{r}
ggplot(smaller, aes(cut, carat)) +
  geom_boxplot()
```

Diamond size varies across cut grades. Fair-cut diamons tend to have a wider
range of carat weights and include more large diamonds. Ideal and Premium cuts also show
considerable variation, although most diamons in every cut category are less
than one carat. Overall, cut quality appears to be associated with differences in diamond size.

## Diamond size by colour

```{r}
ggplot(smaller, aes(color, carat)) +
  geom_boxplot()
```

The distributions of diamond sizes are fairly similar across colour grades. Median carat weights
do not differ substantially, suggesting that colour and size are only weakly related. However, some color categories
contain a wider spread of diamond sizes than others.

## Diamond size by clarity

```{r}
ggplot(smaller, aes(clarity, carat)) +
  geom_boxplot()
```

Diamond size varies somewhat by clarity grade. Lower-clarity diamons tend to include more large stones,
while the highest clarity grades generally contain smaller diamonds on average. This pattern may reflect
the rarity and higher cost of diamons that are both large and exceptionally clear.

## Summary

Diamond size differs across cut, colour, and clarity categories, although the stength of these relationships varies.
Cut appears to have the strongest association with carat weight, while colour shows relatively little variation.
Clarity displays a moderate relationship with size, with larger diamons appearing more frequently among lower clarity grades.


---
27.3.1 Exercise 3: Check that you can run it, then add text after the frequency polygon that describes its most striking features.
---
---
it is characterized by an overall right skew with sharp, highly asymmetric spikes clustered
at standard benchmark values (0.3, 0.5, 0.7, and 1.0 carats), illustrating a strong market preference
for cutting diamonds that meet or slightly exceed these specific weight thresholds
---

---
27.4.7 Exercise 1: Add a section that explores how diamond sizes vary by cut, colour, and clarity. Assume you are writing a report for someone who does not know R, and instead of setting echo = FALSE on each chunk, set a global option.
---
---
solved above
---