---
title: "Predicting Daily Citibike Trips"
author: "Alexandru Soroiu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, message=FALSE}
library(tidyverse)
library(modelr)
```

## Load data

```{r}
trips <- read_tsv("trips_per_day.tsv")
```

## Train/test split

Hold out 10% of the data as a final test set. Use the remaining 90% for
training and validation.

```{r}
set.seed(50)

n <- nrow(trips)

# Test set (10%)
test_idx <- sample(1:n, floor(0.1 * n))
trips_test <- trips[test_idx, ]

# Remaining 90%
trips_train_validate <- trips[-test_idx, ]

# Split remaining into train and validation
n_train_validate <- nrow(trips_train_validate)
train_idx <- sample(1:n_train_validate, floor(0.8 * n_train_validate))

trips_train <- trips_train_validate[train_idx, ]
trips_validate <- trips_train_validate[-train_idx, ]
```

## Baseline model: temperature only

Start with the polynomial model from the complexity control notebook.

```{r}
# TODO: fit a linear model using poly(tmin, degree) on training data
K <- 1:11
train_error <- c()
validate_error <- c()
for (k in K) {
    # fit model on the training data
    model <- lm(num_trips ~ poly(tmin, k, raw = TRUE), data = trips_train)
    
    # compute training RMSE
    train_error[k] <- sqrt(mean((predict(model, trips_train) - trips_train$num_trips)^2))
    
    # compute validation RMSE
    validate_error[k] <- sqrt(mean((predict(model, trips_validate) - trips_validate$num_trips)^2))
}
# TODO: try different polynomial degrees and evaluate RMSE on validation data
```

``` {r}
plot_data <- data.frame(K, train_error, validate_error) %>%
  pivot_longer(cols = c(train_error, validate_error), names_to = "split", values_to = "error")

ggplot(plot_data, aes(x = K, y = error, color = split)) +
  geom_line() +
  scale_x_continuous(breaks = K) +
  labs (x = "Polynomial Degree", y = "RMSE", color = "Dataset")
```

```{r}
best_degree <- K[which.min(validate_error)]

best_degree
```

## Feature engineering

Add features that might help: day of week, month, precipitation, etc.
Remember: only use information available *before* the day in question.

```{r}
trips_features <- trips %>%
    arrange(ymd) %>%
    mutate(
        log_prcp = log(prcp + 1),
        heavy_rain = prcp > 1,
        very_heavy_rain = prcp > 3,
        day_of_week = wday(ymd, label = TRUE),
        month = factor(month(ymd)),
        weekend = day_of_week %in% c("Sat", "Sun"),
        temp_range = tmax - tmin,
        rainy = prcp > 0,
        cold = tmin < 0,
        lag_1_trips = lag(num_trips, 1),
        lag_7_trips = lag(num_trips, 7),
        avg_7_trips = slider::slide_dbl(
            lag(num_trips),
            mean,
            .before = 6,
            .complete = TRUE
        )
    ) %>% 
    drop_na()
```

## New Model with Added Features

``` {r}
set.seed(50)
n <- nrow(trips_features)

test_idx  <- sample(1:n, floor(0.1 * n))
trips_test <- trips_features[test_idx, ]

trips_train_validate <- trips_features[-test_idx, ]

n_train_validate <- nrow(trips_train_validate)
train_idx <- sample(1:n_train_validate, floor(0.8 * n_train_validate))

trips_train <- trips_train_validate[train_idx, ]
trips_validate <- trips_train_validate[-train_idx, ]
```

## Model comparison

Try different models and compare their validation performance.

```{r}
rmse <- function(model, data) {
    sqrt(mean((predict(model, data) - data$num_trips)^2))
}

model_1 <- lm(num_trips ~ poly(tmin, 2, raw = TRUE), data = trips_train)

model_2 <- lm(num_trips ~ poly(tmin, 2, raw = TRUE) + log_prcp + rainy, data = trips_train)

model_3 <- lm(num_trips ~ poly(tmin, 2, raw = TRUE) + log_prcp + very_heavy_rain + weekend + month, data = trips_train)

model_4 <- lm(num_trips ~ poly(tmin, 2, raw = TRUE) + log_prcp + very_heavy_rain + weekend + month + lag_1_trips, data = trips_train)

model_5 <- lm(num_trips ~ poly(tmin, 2, raw = TRUE) + log_prcp + very_heavy_rain + weekend + month + lag_1_trips + avg_7_trips, data = trips_train)

model_6 <- lm(num_trips ~ poly(tmin, 2, raw = TRUE) * weekend + log_prcp + very_heavy_rain + month + lag_1_trips + avg_7_trips, data = trips_train)

model_7 <- lm(log(num_trips) ~ poly(tmin, 2, raw = TRUE) + log_prcp + very_heavy_rain + weekend + month + lag_1_trips + avg_7_trips, data = trips_train)

rmse_log <- function(model, data) {
    preds <- exp(predict(model, data))
    sqrt(mean((preds - data$num_trips)^2))
}

model_results <- tibble(
    model = c(
        "Temp only",
        "Temp + rain",
        "Weather + calendar",
        "Weather + calendar + lag 1",
        "Weather + calendar + lag 1 + avg 7",
        "Interaction model",
        "Log outcome model"
    ),
    validation_rmse = c(
        rmse(model_1, trips_validate),
        rmse(model_2, trips_validate),
        rmse(model_3, trips_validate),
        rmse(model_4, trips_validate),
        rmse(model_5, trips_validate),
        rmse(model_6, trips_validate),
        rmse_log(model_7, trips_validate)
    )
)

model_results %>% 
    arrange(validation_rmse)
```

## Final model

Pick your best model and plot the results.

```{r}
final_model <- lm(
    num_trips ~ poly(tmin, 2, raw = TRUE) * weekend +
    log_prcp + very_heavy_rain + month +
    lag_1_trips + avg_7_trips,
    data = trips_train_validate
)

test_predictions <- predict(final_model, trips_test)

test_rmse <- sqrt(
    mean((test_predictions - trips_test$num_trips)^2)
)
```

## Plot the Final Model
``` {r}
# 1st plot
plot_data <- trips_test %>%
  mutate(predicted_trips = test_predictions)

ggplot(plot_data, aes(x = ymd)) +
  geom_point(aes(y = num_trips), alpha = 0.5) +
  geom_line(aes(y = predicted_trips), color = "red") +
  labs(
    x = "Date",
    y = "Number of Trips",
    title = "Actual and Predicted Daily Citibike Trips"
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

# 2nd plot
plot_data <- trips_test %>% 
    mutate(predicted_trips = test_predictions)

ggplot(plot_data, aes(x = predicted_trips, y = num_trips)) +
    geom_point(alpha = 0.5) +
    geom_abline(
        slope = 1,
        intercept = 0,
        color = "red"
    ) +
    labs(
        x = "Predicted Trips",
        y = "Actual Trips",
        title = "Predicted vs Actual Daily Citibike Trips"
    ) +
    theme_classic() +
    theme(plot.title = element_text(hjust = 0.5))
```


## Save model

```{r}
save(final_model, file = "final_model.RData")
```

## Test set evaluation

**Do this only once, at the very end!**

```{r}
load("final_model.RData")

test_rmse
# TODO: load your saved model and predict on trips_test
# TODO: compute RMSE on test set
# TODO: how do you think this model will perform on future data?
```
The final model had a test RMSE of `r round(test_rmse, 2)`. I expect the model to perform similarly on future data,
although it may perform worse on unusual days such as holidays, extreme weather days, or days with unusually low ridership.
Since the test set was held out until the end, this RMSE is my best estimate of future prediction performance.